Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-15 paper.tex
Update
Remove hyperref
It turns the page into a horrible looking mess.

% \section{Contact:} \href{jerome.kelleher@bdi.ox.ac.uk}{jerome.kelleher@bdi.ox.ac.uk}

@@

% Abstract: ~ 150 words.  Concise summary of the results.

% Key points in abstract.
% 1) Coalescent simulation is a vital tool
% 2) Coalescent with recombination is hard
% 3) Large sample sizes are crucial today
% 4) Msprime has lots of features and uses a community development model

@@

% Keywords: For software, include the relevant models, algorithms and language;
\textbf{Keywords:} Coalescent simulation, Python

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction: Limit of 500 words. The introduction should describe the
% significance of the software or resource presented, discuss novelty, provide an
% overview of the software or resource, and describe how researchers can access it.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Methods: For software, the authors should describe key elements of the algorithms
% implemented, and are expected to provide both the source code as well as
% examples illustrating the use of the software.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Methods}

@@

%introduction
Gene conversion is a unidirectional transfer of a short segment of genetic material,
for example between homologous chromosomes~\citep{Chen2007}.
%description of ancestral process
Just as recombination, gene conversion events split the ancestral material
for some lineage, creating two independent lineages.
In contrast to the previously described recombination process, where the ancestral
material to the left and right of a recombination breakpoint follows two
different ancestral lineages, gene conversion events distribute the genetic material
of a short segment to one ancestral lineage and the part to the left and right of this
segment to another lineage.
%Franz Baumdicker: Not sure if this next sentence is necessary,
% but it might be of interest for some people?
The resulting distribution of ancestral material after a gene conversion event
equals the distribution after two recombination events at the left and right position
of the gene conversion tract, followed by a common ancestor event.
%Impact on LD
However, gene conversions impact much shorter segments than crossover recombinations,
typically below 1kb, and thus affects linkage pattern differently than
recombination~\citep{Korunes2017}.

% other simulators
Hudson's \ms\ optionally includes gene conversion and so does
MaCS~\citep{chen2009fast}, while scrm~\citep{staab2015scrm} does not
support gene conversions.
As bacterial recombination can be modeled analogously to eukaryotic allelic
gene conversion, simulators for microbial evolution often include gene conversion
instead of recombination.
This includes the backward simulator SimBac~\citep{brown2016simbac},
the SMC based approach used in FastSimBac~\citep{DeMaio2017},
and forward in time tools within the SLiM framework~\citep{Cury2020}.

@@

\begin{figure}
\includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf.pdf}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are chosen to mimic the simulation
of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
gene conversion rate was set to $N_e g = \gamma = 0.015$ and
the mean gene conversion tract length was set to $L = 500$, which is close to
estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
with mean tract length $L = 542$~\citep{Lapierre2016}.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Results and Discussion: For software, the authors are expected to present a
% few examples that demonstrate the use of the software and benchmarks. Ideally,
% the examples should be simple. We encourage the use of boxes with scripts
% supported by a narrative. Benchmarks carried under more realistic situations
% than the one presented in the examples are also encouraged.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Results and Discussion}

\subsection*{Discussion}
% Simulation is useful, and it's really easy for a single locus.
% This has lead to a culture of developing lots of simulators
% rather than focusing on one
Simulation has played a key role in the development of coalescent theory
and its use in understanding observed patterns of genetic variation.
The basic model can be simulated for $n$ samples in $O(n)$ time
and is straightforward to implement---for example, Hudson provided a fully
functional and highly efficient simulator as a C source code listing in an
appendix to the seminal review paper~\citep{hudson1990gene}. Because of
this ease, the dominant approach to producing simulators for extensions
to this basic model has been to develop them independently. This is
often seen as a learning opportunity, since those developing extensions
to the coalescent must have an excellent understanding of the basic
model in order to develop extensions to the coalescent.
The coalescent with recombination, however, is \emph{not} straightforward
to implement, and numerous different approaches have been
proposed over the
years~\citep{hudson1983properties,griffiths1997ancestral,wiuf1999recombination,
mcvean2005approximating}. It has only recently become clear that it is
possible to implement the model efficiently, and only through the use
of sophisticated data structures and
intricate algorithms~\citep{kelleher2016efficient}.

@@

\label{app-likelihoods}
\section{Likelihood calculations}

@@

\label{app-multiple-mergers}
\section{Multiple merger coalescent model}

>>>>

% \section*{Contact:} \href{jerome.kelleher@bdi.ox.ac.uk}{jerome.kelleher@bdi.ox.ac.uk}

@@

@@

\textbf{Keywords:} Coalescent simulation, Python

@@

\section*{Results}

@@

%introduction
Gene conversion is a unidirectional transfer of a short segment of genetic material,
for example between homologous chromosomes~\citep{chen2007gene}.
%description of ancestral process
Just as recombination, gene conversion events split the ancestral material
for some lineage, creating two independent lineages.
In contrast to the previously described recombination process, where the ancestral
material to the left and right of a recombination breakpoint follows two
different ancestral lineages, gene conversion events distribute the genetic material
of a short segment to one ancestral lineage and the part to the left and right of this
segment to another lineage.
%Franz Baumdicker: Not sure if this next sentence is necessary,
% but it might be of interest for some people?
The resulting distribution of ancestral material after a gene conversion event
equals the distribution after two recombination events at the left and right position
of the gene conversion tract, followed by a common ancestor event.
%Impact on LD
However, gene conversions impact much shorter segments than crossover recombinations,
typically below 1kb, and thus affects linkage pattern differently than
recombination~\citep{korunes2017gene}.

% other simulators
Hudson's \ms\ optionally includes gene conversion and so does
MaCS~\citep{chen2009fast}, while scrm~\citep{staab2015scrm} does not
support gene conversions.
As bacterial recombination can be modeled analogously to eukaryotic allelic
gene conversion, simulators for microbial evolution often include gene conversion
instead of recombination.
This includes the backward simulator SimBac~\citep{brown2016simbac},
the SMC based approach used in FastSimBac~\citep{demaio2017the},
and forward in time tools within the SLiM framework~\citep{cury2020simulation}.

@@

\begin{figure}
% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf.pdf}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are chosen to mimic the simulation
of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
gene conversion rate was set to $N_e g = \gamma = 0.015$ and
the mean gene conversion tract length was set to $L = 500$, which is close to
estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
with mean tract length $L = 542$~\citep{lapierre2016the}.}
\end{figure}

@@

\section*{Discussion}
% Simulation is useful, and it's really easy for a single locus.
% This has lead to a culture of developing lots of simulators
% rather than focusing on one
Simulation has played a key role in the development of coalescent theory
and its use in understanding observed patterns of genetic variation.
The basic model can be simulated for $n$ samples in $O(n)$ time
and is straightforward to implement---for example, Hudson provided a fully
functional and highly efficient simulator as a C source code listing in an
appendix to the seminal review paper~\citep{hudson1990gene}. Because of
this ease, the dominant approach to producing simulators for extensions
to this basic model has been to develop them independently. This is
often seen as a learning opportunity, since those developing extensions
to the coalescent must have an excellent understanding of the basic
model in order to develop extensions to the coalescent.
The coalescent with recombination, however, is \emph{not} straightforward
to implement, and numerous different approaches have been
proposed over the
years~\citep{hudson1983properties,griffiths1997ancestral,wiuf1999recombination,
mcvean2005approximating}. It has only recently become clear that it is
possible to implement the model efficiently, and only through the use
of sophisticated data structures and
intricate algorithms~\citep{kelleher2016efficient}.

@@

\label{app-likelihoods}
\section*{Likelihood calculations}

@@

\label{app-multiple-mergers}
\section*{Multiple merger coalescent model}
